Demographic Shifts in Toronto's Homeless Population: Evaluating and Reassessing City Solutions

### 2024-09-27 paper/paper.qmd
Updated code style using styler

We dropped the only missing row corresponding to July 2024. We then selected our columns of interest, namely, the total number of homeless people every month (**actively_homeless**), as well as **refugee** and **chronic status** in the column population. In addition, we categorized age groups based on their suitability for work: transforming ages under 24 for the **early career stage**, ages between 25 and 54 for **working age**, and ages above 55 for **pre-retirement and retirement age**. The classification of ages 25-54 as working age is aligned with the Canadian government's survey, which defines individuals aged 25-54 as core-aged workers @statcan2024. Finally, we replaced the monthly data with averaged quarterly data for better analysis and visualization over a prolonged period. A sample of cleaned data can be seen in @tbl-cleandata, and summary statistics in @tbl-summary.

@@

# Give the summary table
cleaned_data |>
  slice(1:5) |>
  kable(
    col.names = c("Time (Quaters)", "Subdivision", "Total people",
                  "Early Career Stage", "Working Age", "Pre-retire and Retire"),
    digits = 0)
```

@@

# Display the summary statistics using kable
summary_statistics |>
  kable(
    col.names = c("Subdivision", "Mean Total People", 
                  "Mean Early Career Stage", 
                  "Mean Working Age", 
                  "Mean Pre-retire and Retire"
  ),
    digits = 2
  )

```

@@

# Filter the data for 'All Population' group
all_population_data <- subset(cleaned_data, population_group == 'All Population')

# Create the bar graph
ggplot(all_population_data, aes(x = Quarter, y = actively_homeless)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(x = "Quarters of a year",
       y = "Total number of homeless people", 
       title = "Total homeless population in toronto, measured quarterly") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

@@

# Create a bar graph
ggplot(plot_data, aes(x = Quarter, y = Proportion)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Quarters of a year", y = "Proportion",
       title = "Proportion of chronic homeless population to toal homeless\npopulation in Toronto, measured quarterly") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
### Age trend
@fig-WorkingAge displays the trend of the proportion of different working age statuses among the actively homeless population over the quarters of each year. Here, the working age statuses are categorized into three groups: Early Career Stage, Working Age, and Pre-Retirement and Retirement Age. Even though the trend is not dramatic, We can see from the graph that the proportion of the Early Career Stage population as well as pre-retire and retired has decreased (from 29% and 17% separately in the first quarter of 2018 to 24.5% and 14.5% in the second quarter in 2024). On the other hand, the working age proportion increased to 61% in 2024Q2 from 54% in 2018Q1. This indicates the growing population who are in the working age cannot afford their housing. 
```{r}
#| label: fig-WorkingAge
#| echo: false
#| fig-cap: ""

# Filter the data for 'All Population' group
all_population_data <- subset(cleaned_data, population_group == 'All Population')

# Calculate the proportion of each working age status
all_population_data <- all_population_data %>%
  mutate(total = Early_Career_Stage + Working_Age + Pre_Retirement_and_Retirement_Age,
         early_career_proportion = Early_Career_Stage / total,
         working_age_proportion = Working_Age / total,
         pre_retirement_proportion = Pre_Retirement_and_Retirement_Age / total)

@@

# Rename age status categories for better readability
proportion_data_long$Age_Status <- recode(proportion_data_long$Age_Status,
                                          early_career_proportion = "Early Career Stage",
                                          working_age_proportion = "Working Age",
                                          pre_retirement_proportion = "Pre-Retirement and Retirement Age")

# Create the stacked bar graph
ggplot(proportion_data_long, aes(x = Quarter, y = Proportion, fill = Age_Status)) +
  geom_bar(stat = "identity") +
  labs(x = "Quarters of a year",
       y = "Proportion",
       title = "Proportion of working age population to to toal homeless\npopulation in Toronto, measured quarterly") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 6)) +
  scale_fill_manual(values = c("Early Career Stage" = "blue", 
                               "Working Age" = "pink", 
                               "Pre-Retirement and Retirement Age" = "green"))
```

@@

# Create a bar graph
ggplot(plot_data, aes(x = Quarter, y = Proportion)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Quarters", y = "Proportion",
       title = "Proportion of Chronic population to to to toal homeless\npopulation in Toronto, measured quarterly") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

>>>>

We dropped the only missing row corresponding to July 2024. We then selected our columns of interest, namely, the total number of homeless people every month (**actively_homeless**), as well as **refugee** and **chronic status** in the column population. In addition, we categorized age groups based on their suitability for work: transforming ages under 24 for the **early career stage**, ages between 25 and 54 for **working age**, and ages above 55 for **pre-retirement and retirement age**. The classification of ages 25-54 as working age is aligned with the Canadian government's survey, which defines individuals aged 25-54 as core-aged workers [@statcan2024]. Finally, we replaced the monthly data with averaged quarterly data for better analysis and visualization over a prolonged period. A sample of cleaned data can be seen in @tbl-cleandata, and summary statistics in @tbl-summary.

@@

# Give the summary table
cleaned_data |>
  slice(1:5) |>
  kable(
    col.names = c(
      "Time (Quaters)", "Subdivision", "Total people",
      "Early Career Stage", "Working Age", "Pre-retire and Retire"
    ),
    digits = 0
  )
```

@@

# Display the summary statistics using kable
summary_statistics |>
  kable(
    col.names = c(
      "Subdivision", "Mean Total People",
      "Mean Early Career Stage",
      "Mean Working Age",
      "Mean Pre-retire and Retire"
    ),
    digits = 2
  )
```

@@

# Filter the data for 'All Population' group
all_population_data <- subset(cleaned_data, population_group == "All Population")

# Create the bar graph
ggplot(all_population_data, aes(x = Quarter, y = actively_homeless)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    x = "Quarters of a year",
    y = "Total number of homeless people",
    title = "Total homeless population in toronto, measured quarterly"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

# Create a bar graph
ggplot(plot_data, aes(x = Quarter, y = Proportion)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(
    x = "Quarters of a year", y = "Proportion",
    title = "Proportion of chronic homeless population to toal homeless\npopulation in Toronto, measured quarterly"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
### Age trend
@fig-WorkingAge displays the trend of the proportion of different working age statuses among the actively homeless population over the quarters of each year. Here, the working age statuses are categorized into three groups: Early Career Stage, Working Age, and Pre-Retirement and Retirement Age. Even though the trend is not dramatic, We can see from the graph that the proportion of the Early Career Stage population as well as pre-retire and retired has decreased (from 29% and 17% separately in the first quarter of 2018 to 24.5% and 14.5% in the second quarter in 2024). On the other hand, the working age proportion increased to 61% in 2024Q2 from 54% in 2018Q1. This indicates the growing population who are in the working age cannot afford their housing. 
```{r}
#| label: fig-WorkingAge
#| echo: false
#| fig-cap: ""

# Filter the data for 'All Population' group
all_population_data <- subset(cleaned_data, population_group == "All Population")

# Calculate the proportion of each working age status
all_population_data <- all_population_data %>%
  mutate(
    total = Early_Career_Stage + Working_Age + Pre_Retirement_and_Retirement_Age,
    early_career_proportion = Early_Career_Stage / total,
    working_age_proportion = Working_Age / total,
    pre_retirement_proportion = Pre_Retirement_and_Retirement_Age / total
  )

@@

# Rename age status categories for better readability
proportion_data_long$Age_Status <- recode(proportion_data_long$Age_Status,
  early_career_proportion = "Early Career Stage",
  working_age_proportion = "Working Age",
  pre_retirement_proportion = "Pre-Retirement and Retirement Age"
)

# Create the stacked bar graph
ggplot(proportion_data_long, aes(x = Quarter, y = Proportion, fill = Age_Status)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Quarters of a year",
    y = "Proportion",
    title = "Proportion of working age population to to toal homeless\npopulation in Toronto, measured quarterly"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 6)
  ) +
  scale_fill_manual(values = c(
    "Early Career Stage" = "blue",
    "Working Age" = "pink",
    "Pre-Retirement and Retirement Age" = "green"
  ))
```

@@

# Create a bar graph
ggplot(plot_data, aes(x = Quarter, y = Proportion)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(
    x = "Quarters", y = "Proportion",
    title = "Proportion of Chronic population to to to toal homeless\npopulation in Toronto, measured quarterly"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
